---
title: Applied Statistical Methods - Notebook 4
author: Peter von Rohr
date: '2023-03-20'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Sum Contrasts
Use the following dataset on `Body Weight` and `Breed` of beef cattle animals. The data is available from

```{r echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_breed.csv"
s_ex04p01_data_path
```

Fit a fixed linear model with `Body Weight` as response and `Breed` as predictor variable. Use the `sum` contrasts for reporting the different effects in the model. Validate the estimates by computing the estimates based on a solution of the least squares normal equations. 

### Your Solution

* Read the dataset given by the above link
* Change contrasts for the factor `Breed` to `sum` contrasts
* Fit linear model with `Body Weight` as response and `Breed` as predictor
* Compute solutions of least squares normal equation
* Use contrasts matrix to determine the estimable functions
* Compare values of estimable functions to output of `lm()`





## Problem 2: Custom Contrasts
Use the dataset from Problem 1 and use your own contrasts. Your new contrasts should compute the intercept estimate as is done in the `sum` contrasts. The `Breed` effects should be computed the same way as is done in the `treatment` contrast.

### Your Solution

* Construct the matrix of estimable functions and of the linear combination for the intercept
* Compute the contrasts matrix
* Fit the linear model using the custom contrasts matrix







```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
