---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 4
author: Peter von Rohr
date: 2023-03-20
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Sum Contrasts
Use the following dataset on `Body Weight` and `Breed` of beef cattle animals. The data is available from

```{r echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_breed.csv"
s_ex04p01_data_path
```

Fit a fixed linear model with `Body Weight` as response and `Breed` as predictor variable. Use the `sum` contrasts for reporting the different effects in the model. Validate the estimates by computing the estimates based on a solution of the least squares normal equations. 

<!-- your-solution-start
### Your Solution

* Read the dataset given by the above link
* Change contrasts for the factor `Breed` to `sum` contrasts
* Fit linear model with `Body Weight` as response and `Breed` as predictor
* Compute solutions of least squares normal equation
* Use contrasts matrix to determine the estimable functions
* Compare values of estimable functions to output of `lm()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


## Problem 2: Custom Contrasts
Use the dataset from Problem 1 and use your own contrasts. Your new contrasts should compute the intercept estimate as is done in the `sum` contrasts. The `Breed` effects should be computed the same way as is done in the `treatment` contrast.

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


## Problem 3

<!-- your-solution-start
### Your Solution

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->
 
