---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 4
author: Peter von Rohr
date: 2023-03-20
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Sum Contrasts
Use the following dataset on `Body Weight` and `Breed` of beef cattle animals. The data is available from

```{r echo=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_breed.csv"
s_ex04p01_data_path
```

Fit a fixed linear model with `Body Weight` as response and `Breed` as predictor variable. Use the `sum` contrasts for reporting the different effects in the model. Validate the estimates by computing the estimates based on a solution of the least squares normal equations. 

<!-- your-solution-start
### Your Solution

* Read the dataset given by the above link
* Change contrasts for the factor `Breed` to `sum` contrasts
* Fit linear model with `Body Weight` as response and `Breed` as predictor
* Compute solutions of least squares normal equation
* Use contrasts matrix to determine the estimable functions
* Compare values of estimable functions to output of `lm()`

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

#### Read the dataset

```{r, message=FALSE, warning=FALSE}
s_ex04p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_breed.csv"
tbl_bw_br <- readr::read_delim(file = s_ex04p01_data_path, delim = ",")
```


#### Change contrasts and fit linear model
The type of contrasts can directly be specified when fitting the linear model. For more information see the help function of `contrasts`

```{r}
lm_bw_br_con_sum <- lm(`Body Weight` ~ Breed, 
                       data = tbl_bw_br,
                       contrasts = list(Breed = "contr.sum"))
(smry_lm_bw_br_con_sum <- summary(lm_bw_br_con_sum))
```


#### Solutions of Least Squares Normal Equations

```{r}
mat_X <- model.matrix(lm(`Body Weight` ~ 0 + Breed, data = tbl_bw_br))
attr(mat_X, "assign") <- NULL
attr(mat_X, "contrasts") <- NULL
colnames(mat_X) <- NULL
mat_X <- cbind(matrix(rep(1,nrow(mat_X)), ncol = 1), mat_X)
mat_xtx <- crossprod(mat_X)
mat_xtx_ginv <- MASS::ginv(mat_xtx)
mat_xty <- crossprod(mat_X, tbl_bw_br$`Body Weight`)
mat_b_sol <- crossprod(mat_xtx_ginv, mat_xty)
mat_b_sol
```


#### Contrasts Matrix for Sum Contrasts
From the contrasts matrix, we get the matrix of estimable functions.

```{r}
fac_breed <- as.factor(tbl_bw_br$Breed)
contr_mat_breed_sum <- contrasts(C(fac_breed, sum))
contr_mat_breed_sum <- cbind(matrix(rep(1,nrow(contr_mat_breed_sum)), ncol = 1), contr_mat_breed_sum)
est_mat_breed_sum <- solve(contr_mat_breed_sum)
est_mat_breed_sum
```

The first row of the above matrix `est_mat_breed_sum` shows how the intercept estimate is computed from the observation means. This means that with the sum contrasts, the intercept is the weighted mean of the mean observation for all breeds. Hence, we get

```{r}
tbl_bw_br_an <- dplyr::filter(tbl_bw_br, Breed == "Angus")
tbl_bw_br_li <- dplyr::filter(tbl_bw_br, Breed == "Limousin")
tbl_bw_br_si <- dplyr::filter(tbl_bw_br, Breed == "Simmental")
sum(c(mean(tbl_bw_br_an$`Body Weight`),
      mean(tbl_bw_br_li$`Body Weight`),
      mean(tbl_bw_br_si$`Body Weight`)))/3
```

Comparing that to the result of `lm()` from above, we see that they are equal.

```{r}
smry_lm_bw_br_con_sum$coefficients["(Intercept)","Estimate"]
```

For the effects estimates, we are looking at the second and the third row of the matrix `est_mat_breed_sum`. We are prepending a column of zeroes to the second and the third row of `est_mat_breed_sum`.

```{r}
mat_q_efun <- cbind(matrix(rep(0, (nrow(est_mat_breed_sum)-1)), ncol = 1), est_mat_breed_sum[2:3,])
crossprod(t(mat_q_efun), mat_b_sol) 
```

These values correspond to the effect estimates from `lm()` 

```{r}
smry_lm_bw_br_con_sum$coefficients[2:3,1]
```


<!-- master-solution-end -->


## Problem 2: Custom Contrasts
Use the dataset from Problem 1 and use your own contrasts. Your new contrasts should compute the intercept estimate as is done in the `sum` contrasts. The `Breed` effects should be computed the same way as is done in the `treatment` contrast.

<!-- your-solution-start
### Your Solution

* Construct the matrix of estimable functions and of the linear combination for the intercept
* Compute the contrasts matrix
* Fit the linear model using the custom contrasts matrix

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

#### Read the dataset

```{r, message=FALSE, warning=FALSE}
s_ex04p02_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_breed.csv"
tbl_bw_br <- readr::read_delim(file = s_ex04p02_data_path, delim = ",")
```

#### Matrix of Estimable Functions
The matrix of estimable functions is a combination of the matrices from the sum contrasts and from the treatment contrasts.

```{r}
fact_breed <- as.factor(tbl_bw_br$Breed)
# treatment
mat_cont_treat <- contrasts(C(fact_breed, treatment))
mat_cont_treat <- cbind(matrix(rep(1, nrow(mat_cont_treat)), ncol = 1), mat_cont_treat)
mat_estf_treat <- solve(mat_cont_treat)
# sum
mat_cont_sum <- contrasts(C(fact_breed, sum))
mat_cont_sum <- cbind(matrix(rep(1, nrow(mat_cont_sum)), ncol = 1), mat_cont_sum)
mat_estf_sum <- solve(mat_cont_sum)
# custom
mat_estf_cust <- rbind(mat_estf_sum[1,], mat_estf_treat[2:3,])
mat_cont_cust <- solve(mat_estf_cust)
mat_cont_cust <- mat_cont_cust[,2:3]
mat_cont_cust
```

Using that contrasts matrix in lm leads to 

```{r}
lm_bw_br_con_cust <- lm(`Body Weight` ~ Breed, 
                       data = tbl_bw_br,
                       contrasts = list(Breed = mat_cont_cust))
(smry_lm_bw_br_con_cust <- summary(lm_bw_br_con_cust))
```

<!-- master-solution-end -->


