---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 5
author: Peter von Rohr
date: 2023-04-24
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Interactions
Use the following dataset on `Breed`, `Breast Circumference` and `Body Weight` and fit a fixed linear effects model with `Body Weight` as response and `Breed` and `Breast Circumference` as predictors and include an interaction term between the two predictors. Compute the expected difference in `Body Weight` for two animals which differ in `Breast Circumference by $1cm$ for every `Breed`. 

The dataset is available under

```{r tbl-ex05-p01, echo=FALSE}
s_tbl_ex05_p01_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_flem.csv"
s_tbl_ex05_p01_path
```


<!-- your-solution-start
### Your Solution

* Read the data
* Fit fixed linear effects model for Body Weight using `Breed` and `Breast Circumference` and the interaction with function `lm()`
* From the solution of the `summary()` of the `lm()` result get the regression coefficient and the interaction effects and compute the expected difference in `Body Weight` for all three breeds. 

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data and select the column that are required for fitting the linear model

```{r}
s_tbl_ex05_p01_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_flem.csv"
tbl_bw_bc_br <- readr::read_delim(s_tbl_ex05_p01_path, delim = ",")
tbl_bw_bc_br <- dplyr::select(tbl_bw_bc_br, Animal, `Body Weight`, `Breast Circumference`, Breed)
tbl_bw_bc_br
```


* Fitting the linear model

```{r}
lm_bw_bc_br_int <- lm(`Body Weight` ~ `Breast Circumference` * Breed, data = tbl_bw_bc_br)
smry_lm_bw_bc_br_int <- summary(lm_bw_bc_br_int)
smry_lm_bw_bc_br_int
```

* Expected difference in body weight for the three breeds:

__Angus__: The expected difference in body weight (in kg) of one centimetre increase in breast circumference corresponds to the regression coefficient of `Breast Circumference` and is 

```{r}
smry_lm_bw_bc_br_int$coefficients["`Breast Circumference`", "Estimate"]
```

__Limousin__: Because, for the breed limousin, there is an interaction effect. We have to add the regression coefficient of `Breast Circumference` to the interaction effect `Breast Circumference`:BreedLimousin. From this we get

```{r}
delta_bw_li <- smry_lm_bw_bc_br_int$coefficients["`Breast Circumference`", "Estimate"] +
  smry_lm_bw_bc_br_int$coefficients["`Breast Circumference`:BreedLimousin", "Estimate"]
delta_bw_li
```

__Simmental__: The same as for limousin, we have for simmental

```{r}
delta_bw_si <- smry_lm_bw_bc_br_int$coefficients["`Breast Circumference`", "Estimate"] +
  smry_lm_bw_bc_br_int$coefficients["`Breast Circumference`:BreedSimmental", "Estimate"]
delta_bw_si
```

<!-- master-solution-end -->


## Problem 2: Simulation
```{r, echo=FALSE, message=FALSE, warning=FALSE}
n_b_intercept <- -1070
n_b_slope <- 8.7
n_res_std_error <- 12
n_nr_obs <- 10
n_slope_tol <- 0.1

# take the values for mean and sd for bc from the dataset
s_tbl_bw_bc_br_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_flem.csv"
tbl_bw_bc_br <- readr::read_delim(s_tbl_bw_bc_br_path, delim = ",")
n_mean_bc <- round(mean(tbl_bw_bc_br$`Breast Circumference`), 0)
n_sd_bc <- round(sd(tbl_bw_bc_br$`Breast Circumference`), 2)
```

Use the following values for intercept and regression slope for `Body Weight` on `Breast Circumference` to simulate a dataset of size $N$. What is the number for $N$ that has to be chosen such that the regression analysis of the simulated data gives the same result as the true regression slope. 

The true values are: 

* Intercept: $`r n_b_intercept`$
* Regression slope: $`r n_b_slope`$
* Residual standard error: $`r n_res_std_error`$

### Hints

* Start with $N=`r n_nr_obs`$, simulate a dataset and analyse the data with `lm()`
* If the result (rounded to 1 digits after decimal point) is not the same then double the size of the dataset, hence use, $N=`r 2*n_nr_obs`$
* Continue until you get close to the true value.
* Assume that the random resiudals follow a normal distribution with mean zero and standard devation equal to $`r n_res_std_error`$
* Take breast circumference to be normally distributed with a mean of $`r n_mean_bc`$ and a standard deviation of $`r n_sd_bc`$
* Use a linear regression model with an intercept to model expected body weight based on breast circumference.


<!-- your-solution-start
### Your Solution

* Start with $N=`r n_nr_obs`$, simulate a dataset and analyse the data with `lm()`
* If the result (rounded to 1 digits after decimal point) is not the same then double the size of the dataset, hence use, $N=`r 2*n_nr_obs`$
* Continue until you get close to the true value.


---  your-solution-end -->


<!-- master-solution-start -->

### Solution
We start with $N=`r n_nr_obs`$ and first generate the matrix $X$ which consists of a column of all ones and a column of breast circumference values in centimetre taken from the given normal distribution. Whenever, we generate some random numbers it is important to first set the seed with the function `set.seed()` to which an integer number is passed. This makes sure that when repeating the simulation the same results are generated.

```{r}
set.seed(1234)
vec_bc <- rnorm(n_nr_obs, mean = n_mean_bc, sd = n_sd_bc)
mat_X <- matrix(c(rep(1,n_nr_obs), vec_bc), ncol = 2)
mat_X
```

Together with the given true values of intercept and slope, and randomly generated residuals, observations are simulates.

```{r}
vec_b <- c(n_b_intercept, n_b_slope)
vec_y <- crossprod(t(mat_X), vec_b) + rnorm(n_nr_obs, mean=0, sd=n_res_std_error)
vec_y
```

The simulated data is analysed with a linear regression model

```{r}
tbl_bw_bc_sim <- tibble::tibble(BodyWeight = vec_y, BreastCircumference=vec_bc)
lm_bw_bc_sim <- lm(BodyWeight ~ BreastCircumference, data = tbl_bw_bc_sim)
lm_bw_bc_sim
```

The absolute deviation between the true value of the slope and the estimated slope from the simulated data is 

```{r}
abs(lm_bw_bc_sim$coefficients[["BreastCircumference"]] - n_b_slope)
```

In the following iteration, the size of the dataset is doubled in each iteration round until, the absolute deviation of the estimated slope from the true value becomes smaller than `r n_slope_tol`.

```{r}
n_max_iter <- 10
n_iter_round <- 0
while(abs(lm_bw_bc_sim$coefficients[["BreastCircumference"]] - n_b_slope) > n_slope_tol && 
      n_iter_round < n_max_iter){
  # count number of iterations and determine number of observations
  n_iter_round <- n_iter_round + 1
  n_nr_obs <- 2 * n_nr_obs
  # simulate breast circumference
  vec_bc <- rnorm(n_nr_obs, mean = n_mean_bc, sd = n_sd_bc)
  mat_X <- matrix(c(rep(1,n_nr_obs), vec_bc), ncol = 2)
  # simulate body weight
  vec_y <- crossprod(t(mat_X), vec_b) + rnorm(n_nr_obs, mean=0, sd=n_res_std_error)
  # analyse simulated data
  tbl_bw_bc_sim <- tibble::tibble(BodyWeight = vec_y, BreastCircumference=vec_bc)
  lm_bw_bc_sim <- lm(BodyWeight ~ BreastCircumference, data = tbl_bw_bc_sim)
  # results
  cat(" * Iteration: ", n_iter_round, "\n")
  cat(" * Number of observations: ", n_nr_obs, "\n")
  cat(" * Regression slope: ", lm_bw_bc_sim$coefficients[["BreastCircumference"]], "\n")
  
}
```



<!-- master-solution-end -->


