---
title: Applied Statistical Methods - Notebook 2
author: Peter von Rohr
date: '2023-03-06'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r ex02-p01-setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
# define data directory depending on online status
if (params$isonline){
  s_data_dir <- "https://charlotte-ngs.github.io/asmss2023/data"
} else {
  s_data_dir <- file.path(here::here(), "docs", "data")
}
s_data_path <- file.path(s_data_dir, "asm_bw_mult_reg.csv")
s_data_bw_bc_path <- file.path(s_data_dir, "asm_bw_bc_reg.csv")
# write the default one variable regression, if we are not online
if (!params$isonline){
  tbl_mult <- readr::read_csv(file = s_data_path)
  tbl_bw_bc <- dplyr::select(tbl_mult, Animal, `Breast Circumference`, `Body Weight`)
  if (!fs::file_exists(path = s_data_bw_bc_path))
    readr::write_csv(tbl_bw_bc, file = s_data_bw_bc_path)
}
tbl_bw_bc <- readr::read_csv(file = s_data_bw_bc_path)
n_nr_obs <- nrow(tbl_bw_bc)
```


## Problem 1: Measurement Unit
The measurement unit has an influence on the results of a regression model. This is demonstrated by changing the unit for `Breast Circumference` (BC) from centimeters to meters. 

### Tasks

* Read the original dataset with BC and BW for `r n_nr_obs` animals. The dataset is availble as csv-file under: `r s_data_bw_bc_path`
* Convert all values of BC from centimeters to meters
* Run the regression model using `lm()`
* Compare the results with the results from the original model

### Your Solution

* Use the function `readr::read_csv()` to read the data

```{r}

```

* Divide all values in column `Breast Circumference` by 100

```{r}

```

* Run the regression using `lm()`

```{r}

```

* Comparison of results





## Problem 2: Significance Level
Do the same type of comparison of regression modelling results when changing the measurement unit for the variable HEI in the complete dataset given in

`https://charlotte-ngs.github.io/asmss2023/data/asm_bw_mult_reg.csv`. 

### Tasks

* Run the same regression model as in Problem 1 of Exercise 1
* Convert the measurement unit for the variable HEI from centimeter to meter
* Compare the results of the two regression models with a special focus on the significance level


### Your Solution

* The same regression model as in Problem 1 of Exercise 1

```{r}

```

* Convert the values in column HEI from centimeter to meter

```{r}

```

* Fit again a multiple regression model

```{r}

```

* Compare the results







```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
