---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 5
author: Peter von Rohr
date: 2023-04-24
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Interactions
Use the following dataset on `Breed`, `Breast Circumference` and `Body Weight` and fit a fixed linear effects model with `Body Weight` as response and `Breed` and `Breast Circumference` as predictors and include an interaction term between the two predictors. Compute the expected difference in `Body Weight` for two animals which differ in `Breast Circumference by $1cm$ for every `Breed`. 

The dataset is available under

```{r tbl-ex05-p01, echo=FALSE}
s_tbl_ex05_p01_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_flem.csv"
s_tbl_ex05_p01_path
```


<!-- your-solution-start
### Your Solution

* Read the data
* Fit fixed linear effects model for Body Weight using `Breed` and `Breast Circumference` and the interaction with function `lm()`
* From the solution of the `summary()` of the `lm()` result get the regression coefficient and the interaction effects and compute the expected difference in `Body Weight` for all three breeds. 

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


## Problem 2: Simulation
Use the following values for intercept and regression slope for `Body Weight` on `Breast Circumference` to simulate a dataset of size $N$. What is the number for $N$ that has to be chosen such that the regression analysis of the simulated data gives the same result as the true regression slope. 

The true values are: 

* Intercept: $-1070$
* Regression slope: $8.7$
* Residual standard error: $12$

### Hints

* Start with $N=10$, simulate a dataset and analyse the data with `lm()`
* If the result (rounded to 1 digits after decimal point) is not the same then double the size of the dataset, hence use, $N=20$
* Continue until you get close to the true value.

<!-- your-solution-start
### Your Solution

* Start with $N=10$, simulate a dataset and analyse the data with `lm()`
* If the result (rounded to 1 digits after decimal point) is not the same then double the size of the dataset, hence use, $N=20$
* Continue until you get close to the true value.


---  your-solution-end -->


<!-- master-solution-start -->

### Solution

<!-- master-solution-end -->


