---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 6
author: Peter von Rohr
date: 2023-05-14
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Sire Model
```{r ex10p01-setup, echo=FALSE}
s_ex10_p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"

sigma_u2 <- 9
sigma_s2 <- sigma_u2 / 4
sigma_e2 <- 36
```

Use the following dataset to predict breeding values using a sire model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex10_p01_data_path, "\n")
```

## Hints

* The variance component $\sigma_s^2$ of the sire effect can be assumed to be  $`r sigma_s2`$.
* The variance component $\sigma_e^2$ of the random resiudals is $`r sigma_e2`$.
* Sex is modelled as a fixed effect.
* The sire pedigree relationship can be computed using the `pedigreemm` package.



<!-- your-solution-start
### Your Solution

* Read the data

* Compute the inverse sire relationship matrix

* Setup the mixed model equations

* Solve mixed model equations

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r, message=FALSE, warning=FALSE}
s_ex10_p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"
tbl_ex10_p01 <- readr::read_csv(s_ex10_p01_data_path)
```


* Compute the inverse sire relationship matrix

```{r}
vec_sire <- unique(tbl_ex10_p01$SIRE)
vec_sire <- vec_sire[!is.na(vec_sire)]
n_nr_sire <- length(vec_sire)
ped_sire <- pedigreemm::pedigree(sire = c(NA, NA, 2),
                                 dam = rep(NA, n_nr_sire),
                                 label = as.character(vec_sire))
mat_A_inv_sire <- as.matrix(pedigreemm::getAInv(ped = ped_sire))
mat_A_inv_sire
```


* Setup the mixed model equations. The mixed model equations are 

$$
\left[ 
\begin{array}{cc}
X^TX  &  X^TZ \\
Z^TX  &  Z^TZ + \lambda_s * A_s^{-1}
\end{array}
\right]
\left[ 
\begin{array}{c}
\hat{b} \\
\hat{s}
\end{array}
\right]
= 
\left[ 
\begin{array}{c}
X^Ty \\
Z^Ty
\end{array}
\right]
$$

where $\lambda_s = \sigma_e^2 / \sigma_s^2$. 

The components of the mixed model equations are shown in the following table

```{r, echo=FALSE}
tbl_mme <- tibble::tibble(Component = c("$X$", "$Z$", "$y$", "$\\lambda_s$", "$A_s^{-1}$"),
                          Description = c("Given in the data",
                                          "Given in the data",
                                          "Given in the data",
                                          "Given by variance components",
                                          "Computed above"))
knitr::kable(tbl_mme,
             booktabs = TRUE,
             longtable = TRUE,
             escape = FALSE)
```

The matrix $X$

```{r}
mat_X <- model.matrix(lm(P ~ 0 + SEX, data = tbl_ex10_p01))
attr(mat_X, "assign") <- NULL
attr(mat_X, "contrasts") <- NULL
colnames(mat_X) <- NULL
mat_X
```

The matrix $Z$

```{r}
mat_Z <- model.matrix(lm(P ~ 0 + as.factor(SIRE), data = tbl_ex10_p01))
attr(mat_Z, "assign") <- NULL
attr(mat_Z, "contrasts") <- NULL
colnames(mat_Z) <- NULL
mat_Z
```

The vector $y$

```{r}
vec_y <- tbl_ex10_p01$P
vec_y
```

The mixed model equations are

```{r}
mat_xtx <- crossprod(mat_X)
mat_xtz <- crossprod(mat_X, mat_Z)
mat_ztx <- t(mat_xtz)
lambda_s <- sigma_e2 / sigma_s2
mat_ztz_a_inv_lambda <- crossprod(mat_Z) + lambda_s * mat_A_inv_sire
mat_coef <- rbind(cbind(mat_xtx, mat_xtz), cbind(mat_ztx, mat_ztz_a_inv_lambda))
mat_xty <- crossprod(mat_X, vec_y)
mat_zty <- crossprod(mat_Z, vec_y)
mat_rhs <- rbind(mat_xty, mat_zty)
```


* Solve mixed model equations. The solution is obtained by

```{r}
mat_sol_sire <- solve(mat_coef, mat_rhs)
mat_sol_sire
```

The solution for the fixed effects are 

```{r}
mat_sol_sire[1:2,]
```

The predicted breeding values are

```{r}
mat_sol_sire[3:nrow(mat_sol_sire),]
```


<!-- master-solution-end -->


## Problem 2: Animal Model
```{r ex10p02-setup, echo=FALSE}
s_ex10_p02_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"

sigma_u2 <- 9
sigma_s2 <- sigma_u2 / 4
sigma_e2 <- 36
```

Use the same dataset as in Problem 1 to predict breeding values, but use an animal model instead of a sire model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex10_p02_data_path, "\n")
```

## Hints

* The variance component $\sigma_u^2$ of the breeding value can be assumed to be  $`r sigma_u2`$.
* The variance component $\sigma_e^2$ of the random resiudals is $`r sigma_e2`$.
* Sex is modelled as a fixed effect.
* The numerator relationship matrix can be computed using the `pedigreemm` package.


<!-- your-solution-start
### Your Solution

* Read the data

* Compute the inverse sire relationship matrix

* Setup the mixed model equations

* Solve mixed model equations

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r, message=FALSE, warning=FALSE}
s_ex10_p02_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"
tbl_ex10_p02 <- readr::read_csv(s_ex10_p02_data_path)
```


* Compute the inverse sire relationship matrix

```{r}
ped <- pedigreemm::pedigree(sire = c(rep(NA, 4), tbl_ex10_p02$SIRE),
                                 dam  = c(rep(NA, 4), tbl_ex10_p02$DAM),
                                 label = as.character(c(1:4, tbl_ex10_p02$ID)))
mat_A_inv <- as.matrix(pedigreemm::getAInv(ped = ped))
mat_A_inv
```


* Setup the mixed model equations. The mixed model equations are 

$$
\left[ 
\begin{array}{cc}
X^TX  &  X^TZ \\
Z^TX  &  Z^TZ + \lambda * A^{-1}
\end{array}
\right]
\left[ 
\begin{array}{c}
\hat{b} \\
\hat{u}
\end{array}
\right]
= 
\left[ 
\begin{array}{c}
X^Ty \\
Z^Ty
\end{array}
\right]
$$

where $\lambda = \sigma_e^2 / \sigma_u^2$. 

The components of the mixed model equations are shown in the following table

```{r, echo=FALSE}
tbl_mme <- tibble::tibble(Component = c("$X$", "$Z$", "$y$", "$\\lambda$", "$A^{-1}$"),
                          Description = c("Given in the data",
                                          "Given in the data",
                                          "Given in the data",
                                          "Given by variance components",
                                          "Computed above"))
knitr::kable(tbl_mme,
             booktabs = TRUE,
             longtable = TRUE,
             escape = FALSE)
```

The matrix $X$

```{r}
mat_X <- model.matrix(lm(P ~ 0 + SEX, data = tbl_ex10_p02))
attr(mat_X, "assign") <- NULL
attr(mat_X, "contrasts") <- NULL
colnames(mat_X) <- NULL
mat_X
```

The matrix $Z$

```{r}
# model matrix from data
mat_Z <- model.matrix(lm(P ~ 0 + as.factor(ID), data = tbl_ex10_p02))
attr(mat_Z, "assign") <- NULL
attr(mat_Z, "contrasts") <- NULL
colnames(mat_Z) <- NULL
# add founders
mat_Z <- cbind(matrix(0, nrow = nrow(mat_Z), ncol = 4), mat_Z)
mat_Z
```

The vector $y$

```{r}
vec_y <- tbl_ex10_p02$P
vec_y
```

The mixed model equations are

```{r}
mat_xtx <- crossprod(mat_X)
mat_xtz <- crossprod(mat_X, mat_Z)
mat_ztx <- t(mat_xtz)
lambda <- sigma_e2 / sigma_u2
mat_ztz_a_inv_lambda <- crossprod(mat_Z) + lambda * mat_A_inv
mat_coef <- rbind(cbind(mat_xtx, mat_xtz), cbind(mat_ztx, mat_ztz_a_inv_lambda))
mat_xty <- crossprod(mat_X, vec_y)
mat_zty <- crossprod(mat_Z, vec_y)
mat_rhs <- rbind(mat_xty, mat_zty)
```


* Solve mixed model equations. The solution is obtained by

```{r}
mat_sol <- solve(mat_coef, mat_rhs)
mat_sol
```

The solution for the fixed effects are 

```{r}
mat_sol[1:2,]
```

The predicted breeding values are

```{r}
mat_sol[3:nrow(mat_sol),]
```

<!-- master-solution-end -->


## Problem 3: Model Comparison
Compare the order of the predicted breeding values for the sires from the sire model and from the animal model. 

<!-- your-solution-start
### Your Solution

* Compare both ordered vectors of predicted breeding values

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Sire model

```{r}
order(mat_sol_sire[3:nrow(mat_sol_sire)], decreasing = TRUE)
```

* Animal model

```{r}
order(mat_sol[3:nrow(mat_sol)], decreasing = TRUE)
```

The order of the sires is the same under both models

<!-- master-solution-end -->
 
