---
title: Applied Statistical Methods - Notebook 6
author: Peter von Rohr
date: '2023-05-14'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Sire Model
```{r ex10p01-setup, echo=FALSE}
s_ex10_p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"

sigma_u2 <- 9
sigma_s2 <- sigma_u2 / 4
sigma_e2 <- 36
```

Use the following dataset to predict breeding values using a sire model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex10_p01_data_path, "\n")
```

## Hints

* The variance component $\sigma_s^2$ of the sire effect can be assumed to be  $`r sigma_s2`$.
* The variance component $\sigma_e^2$ of the random resiudals is $`r sigma_e2`$.
* Sex is modelled as a fixed effect.
* The sire pedigree relationship can be computed using the `pedigreemm` package.



### Your Solution

* Read the data

* Compute the inverse sire relationship matrix

* Setup the mixed model equations

* Solve mixed model equations





## Problem 2: Animal Model
```{r ex10p02-setup, echo=FALSE}
s_ex10_p02_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_ped_sim_data.csv"

sigma_u2 <- 9
sigma_s2 <- sigma_u2 / 4
sigma_e2 <- 36
```

Use the same dataset as in Problem 1 to predict breeding values, but use an animal model instead of a sire model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex10_p02_data_path, "\n")
```

## Hints

* The variance component $\sigma_u^2$ of the breeding value can be assumed to be  $`r sigma_u2`$.
* The variance component $\sigma_e^2$ of the random resiudals is $`r sigma_e2`$.
* Sex is modelled as a fixed effect.
* The numerator relationship matrix can be computed using the `pedigreemm` package.


### Your Solution

* Read the data

* Compute the inverse sire relationship matrix

* Setup the mixed model equations

* Solve mixed model equations





## Problem 3: Model Comparison
Compare the order of the predicted breeding values for the sires from the sire model and from the animal model. 

### Your Solution

* Compare both ordered vectors of predicted breeding values



 


```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
