---
title: "R Student Exercise Platform ASMSS-2023"
author: "Peter von Rohr"
date: "2023-02-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Disclaimer
Preparation of R student exercise platform for asmss-2023 is documented.

# General Info

```{r}
course_name <- "asmss2023"
```

```bash
CRSNAME=asmss2023
```


# Course Directory
The course directory is created on 2-htz

```bash
CRSDIR=/home/quagadmin/courses/$CRSNAME
if [[ -f $CRSDIR ]]; then echo " *** ERROR: Invalid course directory: $CRSDIR ...";fi
mkdir -p $CRSDIR
```


# Dockerfile
The dockerfile for `asmss2023` is created on 2-htz

```bash
CRSNAME=asmss2023
CRSDIR=/home/quagadmin/courses/$CRSNAME
cd $CRSDIR
is [[ ! -d dockerfile ]];then mkdir -p dockerfile;fi
cd dockerfile
echo 'FROM rocker/verse' > Dockerfile
echo >> Dockerfile
echo 'LABEL maintainer="peter.vonrohr@usys.ethz.ch"' >> Dockerfile
echo  >> Dockerfile
echo 'RUN apt-get update -y;apt-get upgrade -y' >> Dockerfile
echo 'RUN apt-get install python3-pip -y' >> Dockerfile
echo 'RUN R -e "install.packages(c(\"pedigreemm\", \"kableExtra\", \"AlphaSimR\", dependencies = TRUE))"'  >> Dockerfile
echo 'RUN R -e "remotes::install_github(repo = \"charlotte-ngs/rmdhelp\")"'  >> Dockerfile
echo 'RUN R -e "remotes::install_github(repo = \"pvrqualitasag/rvcetools\")"'  >> Dockerfile
# check
cat Dockerfile

# build docker image
TODAY=$(date +"%Y%m%d")
docker build --pull --no-cache --tag ${CRSNAME}:$TODAY .

```

# Student List
The list of students is exported as xlsx-file from edoz

```{r}
vec_pkg <- c('openxlsx', 'dplyr', 'tidyr')
if (!all(is.element(vec_pkg, installed.packages())))
  install.packages(vec_pkg, dependencies = TRUE)

# find the name of the xlsx student list
s_xlsx_student_list <- list.files(path = "~/Downloads", pattern = "BelegungenLerneinheit")

# read student list
s_bl_path <- file.path('~/Downloads', s_xlsx_student_list)
wb <- openxlsx::read.xlsx(xlsxFile = s_bl_path)
```

List of students with family names, first names and e-mail addresses

```{r}
library(dplyr)
tbl_student_info <- wb %>% 
    select(Familienname, Rufname, Nummer, `E-Mail`) %>%
    mutate(UDName = `E-Mail`) %>%
    tidyr::separate(UDName, c("Username", NA), sep = '@') %>%
    select(Username, Familienname, Rufname, Nummer, `E-Mail`)
tbl_student_info$Username <- tolower(tbl_student_info$Username)
tbl_student_info
```

Write student info to a file

```{r}
student_dir <- file.path(here::here(), 'students')
if (!dir.exists(student_dir)) fs::dir_create(path = student_dir)
student_file_path <- file.path(student_dir, paste0("students_", course_name, ".txt", collapse = ""))
readr::write_csv(tbl_student_info, file = student_file_path, col_names = FALSE)
```

Test student

```{r}
tbl_test_student_info <- tibble::tibble(Username     = c('vrohrp'),
                                        Familienname = c('von Rohr'), 
                                        Rufname      = c('Peter'), 
                                        Nummer       = c('00-000-000') , 
                                        `E-Mail`     = c('peter.vonrohr@usys.ethz.ch'))
test_student_file_path <- file.path(student_dir, paste0("test_students_", course_name, ".txt", collapse = ""))
readr::write_csv(tbl_test_student_info, file = test_student_file_path, col_names = FALSE)

```

## Student Usernames
From the e-mail addresses, we extract usernames as follows

```{r}
(tbl_student_username <- tbl_student_info %>% select(Username))
```

Write the usernames to a file

```{r}
user_file_path <- file.path(student_dir, paste0("student_usernames_", course_name, ".txt", collapse = ""))
readr::write_csv(tbl_student_username, file = user_file_path, col_names = FALSE)
```

The username for the test-student is specified manually and is not extracted from the e-mail. 

```{r}
test_user_file_path <- file.path(student_dir, paste0("test_student_usernames_", course_name, ".txt", collapse = ""))
tbl_test_student_username <- tbl_test_student_info %>% select(Username)
readr::write_csv(tbl_test_student_username, file = test_user_file_path, col_names = FALSE)
```


# File Upload
Upload the created files to the server

```
cd /Users/pvr/Data/Projects/GitHub/charlotte-ngs/${CRSNAME}_gh-root/main/${CRSNAME}
TRGCRSPATH="/home/quagadmin/courses/$CRSNAME"
scp -r students quagadmin@2-htz.quagzws.com:"$TRGCRSPATH"
scp -r bash quagadmin@2-htz.quagzws.com:"$TRGCRSPATH"
```


# Start Containers
On server 2-htz

Check which students already had a lbgfs2022 account and copy home-dir

```bash
CRSNAME=asmss2023
PRVCRS=lbgfs2022
cd /home/quagadmin/courses/$CRSNAME
mkdir -p home
cd home
# test
cat ../students/test_student_usernames_asmss2023.txt | while read s
do
  echo " * Student: $s ..."
  if [[ -d "/home/quagadmin/courses/$PRVCRS/home/$s" ]];then
    echo " * Copy /home/quagadmin/courses/$PRVCRS/home/$s ..."
    cp -r /home/quagadmin/courses/$PRVCRS/home/$s .
  else
    echo " * Create user directory $s ..."
    mkdir -p $s
  fi
done

# real
cat ../students/student_usernames_asmss2023.txt | while read s
do
  echo " * Student: $s ..."
  if [[ -d "/home/quagadmin/courses/$PRVCRS/home/$s" ]];then
    echo " * Copy /home/quagadmin/courses/$PRVCRS/home/$s ..."
    cp -r /home/quagadmin/courses/$PRVCRS/home/$s .
  else
    echo " * Create user directory $s ..."
    mkdir -p $s
  fi
done

```

Copy e-mail sent info

```bash
CRSNAME=asmss2023
PRVCRS=lbgfs2022
cd /home/quagadmin/courses/$CRSNAME
mkdir -p email_sent
cd email_sent
cat ../students/test_student_usernames_asmss2023.txt | while read s
do
  echo " * Student: $s ..."
  if [[ -f /home/quagadmin/courses/${PRVCRS}/email_sent/${s}_email.txt ]];then
    cp /home/quagadmin/courses/${PRVCRS}/email_sent/${s}_email.txt .
  fi
done
cat ../students/student_usernames_asmss2023.txt | while read s
do
  echo " * Student: $s ..."
  if [[ -f /home/quagadmin/courses/${PRVCRS}/email_sent/${s}_email.txt ]];then
    cp /home/quagadmin/courses/${PRVCRS}/email_sent/${s}_email.txt .
  fi
done
```

# Create table with student links

```bash
CRSNAME=asmss2023
cd /home/quagadmin/courses/$CRSNAME
cat students/test_students_${CRSNAME}.txt  | while read s
do
  user=$(echo "$s" | cut -d ',' -f1)
  port=$(cat email_sent/${user}_email.txt | grep Port | cut -d ' ' -f2)
  echo "${user};${port}" >> test_students_rexpf_ports.csv
  sleep 2
done  

cat students/students_${CRSNAME}.txt  | while read s
do
    user=$(echo "$s" | cut -d ',' -f1)
  if [[ -f email_sent/${user}_email.txt ]]; then
    echo " * Known user with previous rexpf instance: $s ..."
    port=$(cat email_sent/${user}_email.txt | grep Port | cut -d ' ' -f2)
    echo "${user};${port}" >> students_rexpf_ports.csv
  else
    echo " * Unknown user $s ..."
  fi  
  sleep 2
done  
```

New users are put on new ports. Search first port that has not been taken by lbgfs2022

```bash
MAXPORT=0
CRSNAME=asmss2023
PRVCRS=lbgfs2022
cd /home/quagadmin/courses/$CRSNAME
for f in $(ls -1 /home/quagadmin/courses/${PRVCRS}/email_sent/*.txt | grep -v 'vrohrp')
do
  port=$(cat $f | grep 'Port' | cut -d ' ' -f2)
  echo " * Extract port: $port from $f ..."
  if [[ $port -gt $MAXPORT ]];then
    MAXPORT=$port
    echo " * MAXPORT: $MAXPORT ..."
  fi
done
echo "$((MAXPORT+1))" > start_port_${CRSNAME}.txt
```

Adding ports for new users

```bash
CRSNAME=asmss2023
cd /home/quagadmin/courses/$CRSNAME
CURPORT=$(cat start_port_${CRSNAME}.txt)
cat students/students_${CRSNAME}.txt  | while read s
do
  user=$(echo "$s" | cut -d ',' -f1)
  if [[ $(grep "$user" students_rexpf_ports.csv | wc -l) -eq 0 ]];then
    echo " * Add new user: $user ..."
    echo "${user};$CURPORT" >> students_rexpf_ports.csv
    CURPORT=$((CURPORT+1))
  else
    echo " * Existing user: $user ..."
  fi
done  
```

Create student info db into csv-file

```bash
CRSNAME=asmss2023
DBINFO=students/test_students_rexpf_db.csv
cd /home/quagadmin/courses/$CRSNAME
if [[ -f $DBINFO ]];then rm $DBINFO;fi
cat test_students_rexpf_ports.csv | while read l
do
  user=$(echo $l | cut -d ';' -f1)
  port=$(echo $l | cut -d ';' -f2)
  echo " * User: $user ..."
  echo " * Port: $port ..."
  # generate new passwd
  passwd=$(tr -dc A-Za-z0-9_ < /dev/urandom | head -c12)
  echo " * Generated pwd: $passwd ..."
  # write info to db-file
  echo "${user},${passwd},$port" >> $DBINFO
  echo
  sleep 2
done  

DBINFO=students/students_rexpf_db.csv
cd /home/quagadmin/courses/$CRSNAME
if [[ -f $DBINFO ]];then rm $DBINFO;fi
cat students_rexpf_ports.csv | while read l
do
  user=$(echo $l | cut -d ';' -f1)
  port=$(echo $l | cut -d ';' -f2)
  echo " * User: $user ..."
  echo " * Port: $port ..."
  # generate new passwd
  passwd=$(tr -dc A-Za-z0-9_ < /dev/urandom | head -c12)
  echo " * Generated pwd: $passwd ..."
  # write info to db-file
  echo "${user},${passwd},$port" >> $DBINFO
  echo
  sleep 2
done  

```

Upload new docker start script

```bash
scp bash/docker_student_cnt_start.sh quagadmin@2-htz.quagzws.com:/home/quagadmin/courses/asmss2023/bash
```

Test starting docker container

```bash
CRSNAME=asmss2023
TODAY=$(date +"%Y%m%d")

cd /home/quagadmin/courses/$CRSNAME
cat students/test_students_rexpf_db.csv | while read l
do
  user=$(echo $l | cut -d ',' -f1)
  passwd=$(echo $l | cut -d ',' -f2)
  port=$(echo $l | cut -d ',' -f3)
  ./bash/docker_student_cnt_start.sh -i ${CRSNAME}:$TODAY -p $port -u $user -w $passwd -v /home/quagadmin/courses/asmss2023/home/$user
  sleep 2
done

CRSNAME=asmss2023
TODAY=$(date +"%Y%m%d")
cd /home/quagadmin/courses/$CRSNAME
cat students/students_rexpf_db.csv | while read l
do
  user=$(echo $l | cut -d ',' -f1)
  passwd=$(echo $l | cut -d ',' -f2)
  port=$(echo $l | cut -d ',' -f3)
  ./bash/docker_student_cnt_start.sh -i ${CRSNAME}:$TODAY -p $port -u $user -w $passwd -v /home/quagadmin/courses/asmss2023/home/$user
  sleep 2
done

```

Stop docker container

```bash
CRSNAME=asmss2023
cd /home/quagadmin/courses/$CRSNAME
cat students/test_students_rexpf_db.csv | while read l
do
  user=$(echo $l | cut -d ',' -f1)
  docker stop ${user}_rstudio
  sleep 2
done

CRSNAME=asmss2023
cd /home/quagadmin/courses/$CRSNAME
cat students/students_rexpf_db.csv | while read l
do
  user=$(echo $l | cut -d ',' -f1)
  docker stop ${user}_rstudio
  sleep 2
done
```


```
CRSNAME=asmss2023
cd /home/quagadmin/courses/$CRSNAME/bash
# test
./docker_student_start.sh -f ../students/test_students_${CRSNAME}.txt -c /home/quagadmin/courses/${CRSNAME}/home -p 10187
# real
./docker_student_start.sh -f ../students/students_${CRSNAME}.txt -c /home/quagadmin/courses/${CRSNAME}/home
```


