---
title: Applied Statistical Methods - Notebook 1
author: Peter von Rohr
date: '2023-02-27'
output: html_notebook
params:
  doctype:
    label: Document Type
    value: solution
    choices:
    - exercise
    - solution
    - notebook
  isonline:
    label: Online (y/n)
    value: true
    choices:
    - true
    - false
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Regression Analysis
The following dataset on body weight and on futher observations on a number of animals is given.

```{r tbl-ex01-p01-data, echo=FALSE, message=FALSE, warning=FALSE}
s_ex01_p01_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_mult_reg.csv"
tbl_ex01_p01_data <- readr::read_csv(file = s_ex01_p01_path)
knitr::kable(tbl_ex01_p01_data,
             booktabs = TRUE,
             longtable = TRUE,
             escape = FALSE)
```

The data can be read from `r s_ex01_p01_path`. The additional columns contain data on body condition score (BCS) and height (HEI).

### Tasks

* Build a regression model of body weight on the other observations using the dataset given above.
* Set up the matrix $\mathbf{X}$ and the vectors $\mathbf{y}$, $\mathbf{b}$ and $\mathbf{e}$.
* Compute estimate for the regression coefficients in the model defined above.


### Your Solution

* Start by building the model which includes the decision of what type of components go into the vector $\mathbf{b}$. Then write down the model in matrix-vector notation.
* Determine all the components of $\mathbf{X}$, $\mathbf{y}$, $\mathbf{b}$ and $\mathbf{e}$.
* Compute the solution for $\hat{\mathbf{b}}





## Problem 2

Use the same dataset as in Problem 1 and verify your results using the function `lm()` in R. 

### Your Solution

* Read the data into a dataframe
* Use the `lm()` function which takes a formula to define the model and a dataframe
* Use the function `summary()` on the result of `lm()` to show the results



 


```{r, echo=FALSE, results='asis'}
cat('\n---\n\n _Latest Changes: ', format(Sys.time(), '%Y-%m-%d %H:%M:%S'), ' (', Sys.info()['user'], ')_\n', sep = '')
```
 
