---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 7
author: Peter von Rohr
date: 2023-05-14
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Marker Effects Model
```{r ex11p01-setup, echo=FALSE}
s_ex11_p01_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_geno_sim_data.csv"

sigma_q2 <- 3
sigma_e2 <- 36
```

Predict genomic breeding values using a marker effects model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex11_p01_data_path, "\n")
```

### Hints

* The variance $\sigma_q^2$ of the marker effect is $`r sigma_q2`$.
* The residual variance $\sigma_e^2$ is $`r sigma_e2`$
* The sex of each animal can be modelled as a fixed effect

<!-- your-solution-start
### Your Solution

* Read the data
* Setup mixed model equations to predict marker effects for all the SNP-loci
* Compute predicted genomic breeding values based on the estimated marker effects

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r, message=FALSE, warning=FALSE}
tbl_ex11_p01 <- readr::read_csv(s_ex11_p01_data_path)
```

* Setup mixed model equations to predict marker effects for all the SNP-loci. The model is given as 

$$y = Xb + Wq + e$$
where $y$ is the vector of observations, $b$ is the vector of fixed effects and $q$ is the vector of random marker effects for each SNP. The matrices $X$ and $W$ are design matrices. The matrix $W$ is special because it contains the genotype encodings. 

From that model the mixed model equations can be specified as

$$ 
\left[ 
\begin{array}{cc}
X^TX  &  X^TW \\
W^TX  &  W^TW + \lambda_q * I
\end{array}
\right]
\left[ 
\begin{array}{c}
\hat{b} \\
\hat{q}
\end{array}
\right]
= 
\left[ 
\begin{array}{c}
X^Ty \\
W^Ty
\end{array}
\right]
$$
with $\lambda_q = \sigma_e^2 / \sigma_q^2$. 

The matrix $X$

```{r}
mat_X <- model.matrix(lm(P ~ 0 + SEX, data = tbl_ex11_p01))
attr(mat_X, "assign") <- NULL
attr(mat_X, "contrasts") <- NULL
mat_X
```

The matrix $W$

```{r}
library(dplyr)
tbl_geno_ex11_p01 <- tbl_ex11_p01 %>%
  select(SNP1:SNP100)
mat_W <- as.matrix(tbl_geno_ex11_p01)
mat_W[,1:10]
```

The vector $y$

```{r}
vec_y <- tbl_ex11_p01$P
vec_y
```

The mixed model equations

```{r}
# coefficient matrix
mat_xtx <- crossprod(mat_X)
mat_xtw <- crossprod(mat_X, mat_W)
mat_wtx <- t(mat_xtw)
lambda_q <- sigma_e2 / sigma_q2
mat_ztz_lambda_I <- crossprod(mat_W) + lambda_q * diag(1, nrow = ncol(mat_W))
mat_coef <- rbind(cbind(mat_xtx, mat_xtw), 
                  cbind(mat_wtx, mat_ztz_lambda_I))
# right hand side
mat_xty <- crossprod(mat_X, vec_y)
mat_wty <- crossprod(mat_W, vec_y)
mat_rhs <- rbind(mat_xty, mat_wty)
# solution
mat_sol <- solve(mat_coef, mat_rhs)
# partition solutions
vec_sol_fix <- mat_sol[1:2,]
vec_sol_marker <- mat_sol[3:nrow(mat_sol),]
```

The solution for the estimates of the fixed effects are: 

```{r}
vec_sol_fix
```

The solutions for the first few marker effects are

```{r}
vec_sol_marker[1:10]
```


* Compute predicted genomic breeding values based on the estimated marker effects. The predicted genomic breeding values are obtained by the matrix-multiplication of matrix $W$ times the vector of the estimated marker effects.

```{r}
mat_mem_gbv <- crossprod(t(mat_W), vec_sol_marker)
mat_mem_gbv
```


<!-- master-solution-end -->


## Problem 2: Breeding Value Based Model
```{r ex11p02-setup, echo=FALSE}
s_ex11_p02_data_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_geno_sim_data.csv"

sigma_q2 <- 3
sigma_u2 <- 3*sigma_q2
sigma_e2 <- 36
```

Use the same dataset as in Problem 1 to predict genomic breeding values based on a breeding-value model. The dataset is available from

```{r, echo=FALSE}
cat(s_ex11_p02_data_path, "\n")
```

### Hints

* The genomic variance $\sigma_u^2$ of the marker effect is $`r sigma_u2`$.
* The residual variance $\sigma_e^2$ is $`r sigma_e2`$
* The sex of each animal can be modelled as a fixed effect
* Use the following function to compute the genomic relationship matrix $G$ based on the matrix of genotypes

```{r}
computeMatGrm <- function(pmatData) {
  matData <- pmatData
  # check the coding, if matData is -1, 0, 1 coded, then add 1 to get to 0, 1, 2 coding
  if (min(matData) < 0) matData <- matData + 1
  # Allele frequencies, column vector of P and sum of frequency products
  freq <- apply(matData, 2, mean) / 2
  P <- 2 * (freq - 0.5)
  sumpq <- sum(freq*(1-freq))
  # Changing the coding from (0,1,2) to (-1,0,1) and subtract matrix P
  Z <- matData - 1 - matrix(P, nrow = nrow(matData), 
                             ncol = ncol(matData), 
                             byrow = TRUE)
  # Z%*%Zt is replaced by tcrossprod(Z)
  return(tcrossprod(Z)/(2*sumpq))
}
```

* If the genomic relationship matrix $G$ which is computed by the function above cannot be inverted, add $0.05 * I$ to $G$ which results in $G^*$ and use $G^*$ as genomic relationship matrix.


<!-- your-solution-start
### Your Solution

* Read the data
* Compute the inverse genomic relationship matrix using the given function for the genomic relationship matrix
* Setup mixed model equations to predict genomic breeding values

---  your-solution-end -->


<!-- master-solution-start -->

### Solution

* Read the data

```{r, message=FALSE, warning=FALSE}
tbl_ex11_p02 <- readr::read_csv(s_ex11_p02_data_path)
```

* Compute the inverse genomic relationship matrix using the given function for the genomic relationship matrix. The genomic relationship matrix $G$ is computed using the above given function with the matrix $W$ from the marker effect model as an argument.

The matrix $W$

```{r, message=FALSE, warning=FALSE}
library(dplyr)
tbl_geno_ex11_p02 <- tbl_ex11_p02 %>%
  select(SNP1:SNP100)
mat_W <- as.matrix(tbl_geno_ex11_p01)
```

The genomic relationship matrix $G$

```{r}
mat_G <- computeMatGrm(pmatData = mat_W)
mat_G
```

We have to check whether $G$ can be inverted. This is done by computing the rank of the matrix

```{r}
Matrix::rankMatrix(mat_G)
```

This shows that matrix $G$ does not have full column rank. Hence we add $0.05 * I$ to get to matrix $G^*$. 

```{r}
mat_G_star <- mat_G + 0.05 * diag(1, nrow = nrow(mat_G))
Matrix::rankMatrix(mat_G_star)
```

Matrix $G^*$ can be used as genomic relationship matrix.

* Setup mixed model equations to predict genomic breeding values. The breeding value based model is given by 

$$ y = Xb + Zu + e$$
where $y$ is the vector of observations, $b$ is the vector of fixed effects and $u$ is the vector of random genomic breeding values. The matrices $X$ and $Z$ are design matrices. 

The mixed model equations are

$$
\left[ 
\begin{array}{cc}
X^TX  &  X^TZ \\
Z^TX  &  Z^TZ + \lambda_g * (G^*)^{-1}
\end{array}
\right]
\left[ 
\begin{array}{c}
\hat{b} \\
\hat{u}
\end{array}
\right]
= 
\left[ 
\begin{array}{c}
X^Ty \\
Z^Ty
\end{array}
\right]
$$

with $\lambda_u = \sigma_e^2 / \sigma_u^2$. 

The matrix $X$

```{r}
mat_X <- model.matrix(lm(P ~ 0 + SEX, data = tbl_ex11_p02))
attr(mat_X, "assign") <- NULL
attr(mat_X, "contrasts") <- NULL
colnames(mat_X) <- NULL
mat_X
```

The matrix $Z$

```{r}
# model matrix from data
mat_Z <- model.matrix(lm(P ~ 0 + as.factor(ID), data = tbl_ex11_p02))
attr(mat_Z, "assign") <- NULL
attr(mat_Z, "contrasts") <- NULL
colnames(mat_Z) <- NULL
mat_Z
```

The vector $y$

```{r}
vec_y <- tbl_ex11_p02$P
vec_y
```

The mixed model equations are

```{r}
# coefficient matrix
mat_xtx <- crossprod(mat_X)
mat_xtz <- crossprod(mat_X, mat_Z)
mat_ztx <- t(mat_xtz)
lambda_u <- sigma_e2 / sigma_u2
mat_ztz_u_inv_lambda <- crossprod(mat_Z) + lambda_u * mat_G_star
mat_coef <- rbind(cbind(mat_xtx, mat_xtz), cbind(mat_ztx, mat_ztz_u_inv_lambda))
# right hand side
mat_xty <- crossprod(mat_X, vec_y)
mat_zty <- crossprod(mat_Z, vec_y)
mat_rhs <- rbind(mat_xty, mat_zty)
# solution
mat_sol <- solve(mat_coef, mat_rhs)
# partition the solution
vec_sol_fix <- mat_sol[1:2,]
vec_sol_gbv <- mat_sol[3:nrow(mat_sol),]
```

The solution for the estimated fixed effects are 

```{r}
vec_sol_fix
```

The predicted genomic breeding values are

```{r}
vec_sol_gbv
```

Comparing order of animals according to predicted genomic breeding values from Problem 1 and Problem 2:

* marker effect model

```{r}
order(mat_mem_gbv[,1], decreasing = TRUE)
```

* breeding value based model

```{r}
order(vec_sol_gbv, decreasing = TRUE)
```



<!-- master-solution-end -->

