---
title: Applied Statistical Methods - `r tools::toTitleCase(params$doctype)` 1
author: Peter von Rohr
date: 2023-02-27
output:
  bookdown::pdf_document2:
    toc: false
    number_sections: false
    keep_md: false
    keep_tex: false
header-includes:
 \usepackage{longtable}
 \usepackage{float}
params:
  doctype:
    label: 'Document Type'
    value: 'solution'
    choices: ['exercise', 'solution', 'notebook']
  isonline:
    label: 'Online (y/n)'
    value: TRUE
    choices: [TRUE, FALSE]
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Problem 1: Regression Analysis
The following dataset on body weight and on futher observations on a number of animals is given.

```{r tbl-ex01-p01-data, echo=FALSE, message=FALSE, warning=FALSE}
s_ex01_p01_path <- "https://charlotte-ngs.github.io/asmss2023/data/asm_bw_mult_reg.csv"
tbl_ex01_p01_data <- readr::read_csv(file = s_ex01_p01_path)
knitr::kable(tbl_ex01_p01_data,
             booktabs = TRUE,
             longtable = TRUE,
             escape = FALSE)
```

The data can be read from `r s_ex01_p01_path`. The additional columns contain data on body condition score (BCS) and height (HEI).

### Tasks

* Build a regression model of body weight on the other observations using the dataset given above.
* Set up the matrix $\mathbf{X}$ and the vectors $\mathbf{y}$, $\mathbf{b}$ and $\mathbf{e}$.
* Compute estimate for the regression coefficients in the model defined above.






## Problem 2

Use the same dataset as in Problem 1 and verify your results using the function `lm()` in R. 



  
