---
title: "ASM Exam Sol Problem 1: Linear Regression"
output: html_notebook
---

# Data Simulation
Descriptive statistics according to 

>"Supplementary Table S2. Summary statistics of milk non-volatile metabolites measured (relative peak area)"

```{r}
# Acetone 
n_mean <- 0.035 
n_sd <- 0.0159 
n_min <- 0.023 
n_max <- 0.102
```



According to Table 4 of https://doi.org/10.3168/jds.2015-10248, we use the data based on 

> "Linear regression between methane yield (g/kg of DMI) and milk nonvolatile metabolites (relative area)"

We use the metabolite with highest $R^2$ which is Acetone

```{r}
# Acetone and regression measures
n_intercept_value <- 25.6
n_intercept_se <- 0.79
n_slope_value <- -43.20
n_slope_se <- 20.664
n_slope_p <- 0.046
n_slope_r2 <- 0.135
```

Characteristics of dataset

```{r}
n_nr_cow <- 12
```

The dataset for `r n_nr_cow` cows is given by

```{r}
# setting the seed by
#rteachtools::get_seed()
set.seed(8968)
tbl_ace_meth <- tibble::tibble(Cow = c(1:n_nr_cow),
                               Acetone = rnorm(n_nr_cow, mean = n_mean, sd = n_sd))

cat("Min Aceton: ", min(tbl_ace_meth$Acetone), " -- ", n_min, "\n")
cat("Max Aceton: ", max(tbl_ace_meth$Acetone), " -- ", n_max, "\n")

# adjust sd to fit given min and max. Not done currently , because sd is getting reduced substantially
b_adjust_sd <- FALSE

# check for minimum and iterate until condition is met
if (b_adjust_sd){
  n_max_iter <- 10
  n_cnt_iter <- 0
  n_cur_sd <- n_sd
  while (min(tbl_ace_meth$Acetone) < n_min || 
         max(tbl_ace_meth$Acetone) > n_max || 
         n_max_iter > n_cnt_iter){
    n_cnt_iter <- n_cnt_iter + 1
    n_cur_sd <- 0.95 * n_cur_sd
    tbl_ace_meth$Acetone <- rnorm(n_nr_cow, mean = n_mean, sd = n_cur_sd)
  }
  cat("Iterations: ", n_cnt_iter, "\n")
  cat("Min Aceton: ", min(tbl_ace_meth$Acetone), " -- ", n_min, "\n")
  cat("Max Aceton: ", max(tbl_ace_meth$Acetone), " -- ", n_max, "\n")
}

```

Generate $CH_4$ data based on given regression

```{r}
n_methan_sd <- 0.38
tbl_ace_meth$Methane <- n_intercept_value + n_slope_value * tbl_ace_meth$Acetone + rnorm(n_nr_cow, mean = 0, sd = n_methan_sd)
tbl_ace_meth_rounded <- tbl_ace_meth
tbl_ace_meth_rounded$Acetone <- round(tbl_ace_meth$Acetone, digits = 4)
tbl_ace_meth_rounded$Methane <- round(tbl_ace_meth$Methane, digits = 2)
tbl_ace_meth_rounded
```

Write the data

```{r}
s_ace_meth_path <- file.path(here::here(), "docs", "data", "asm_exam_p01.csv")
readr::write_delim(tbl_ace_meth_rounded, file = s_ace_meth_path, delim = ",")
```

Regression

```{r}
lm_meth_ace <- lm(Methane ~ Acetone, data = tbl_ace_meth_rounded)
lm_meth_ace
```

Plots


```{r}
library(ggplot2)
ggplot(data = tbl_ace_meth_rounded,
       aes(x = Acetone, y = Methane)) + 
  geom_point() + 
  geom_abline(slope = lm_meth_ace$coefficients[["Acetone"]],
              intercept = lm_meth_ace$coefficients[["(Intercept)"]],
              color = "red") + 
  xlim(c(0, 1.1*max(tbl_ace_meth_rounded$Acetone))) + 
  ylim(c(0.95 * min(tbl_ace_meth_rounded$Methane),  lm_meth_ace$coefficients[["(Intercept)"]]))
         
```

